Inefficient Concessions and Mediation

### 2017-09-12 draft.tex
new proof and exposition around theorem 4

\section{Analysis of benchmark cases}
\label{sec:bench}
When countries are faced with doubts about each other's motives, they have several options of how to proceed. There are three basic types of equilibria that this analysis will examine.\footnote{If the discount factor for the high type ($\delta_h$) exceeds the cutoff found in expression (1), there are many types of equilibria; consider for instance oscillating every other period between Peace and War with a grim trigger War threat. In this paper, we are not interested in these equilibria because they are not as realistic for applications to conflict scenarios. These are also always payoff dominated by at least one of the other equilibria discussed, so are not attractive.} A thorough analysis of the different types of equilibria, as well as all proofs for this section, can be found in the appendix. 
\\
\\
{\bf \emph{Pooling equilibrium}\\}
\vskip-.2in
In the \emph{pooling equilibrium}, both countries do not give concessions in period 0, then always play Fight in every period beginning with period 1. 
Countries continue to doubt, doing nothing to bridge the credibility gap, and thus not cooperating. This is a \emph{pooling equilibrium}, and is likely to happen when countries are in long-set patterns of distrust. Active fighting is not necessary; remaining outside of a state of peace is all that is required.  See Lemma~\ref{lemma:1} and Lemma~\ref{lemma:2} in the appendix for a formal characterization of this type of equilibrium.

@@

The patience threshold to separate through concessions is much higher under these parameters when the concessions carry future material value. In order to calculate it, we need to know the size of the minimum separating concession, which is again calculated from the low-type incentive constraint. The minimum separating concession is $g^1 = \frac{p(1-d_l)(D+T)}{(1-d_l)(1-pT)+(1-p)D} = .\overline{63}$. The minimum patience level required for separation is $\de_h^1 = D(1-p)+1-pT + \frac{p(W-(T+D))}{g^1} = .7\overline{9}.$ Because the specified $\de_h$ is not involved in any of these calculations, we can see that for any $\de_h \in [.5\overline{3},.7\overline{9})$, separating through concessions is possible when concessions have no material value but is not possible when they do. \hfill $\blacksquare$

@@

\begin{theorem}
		Assume the high types are not patient enough to separate without concessions and that concessions have material value that can be destroyed. There are parameter under which the optimal equilibrium is a separating equilibrium in which concessions are inefficient.
	\label{theorem:4}
\end{theorem}
 \emph{Proof}:   Because it is relatively complicated to calculate equilibrium concessions in the general case, a slight simplification of the general model will be made for proof of inefficient concessions. Restrict functional forms of $T(\cdot),W(\cdot)$ and $D(\cdot)$ so that each is a scalar multiplied by its arguments. Also assume that $s_1=s_2=m_1=m_2=1.$ Because of linearity, equilibrium optimization has high types choosing $\alpha=1$ (because a concession would only come from another high type in equilibrium) and low types choosing $\alpha=0$. Note that none of these simplifications is essential for the results to hold but we make them for ease of calculation and demonstration of the results. 
The low type's IC constraint binds so with $e=1$, $g^*$ can be calculated to be $g^*= \frac{p(1-d_l)(D+T)}{(1-d_l)(1-pT)+(1-p)D}$.
Under an equilibrium with purely inefficient concessions $(e=0)$, the minimum separating concession would be $g'=p(T+D)$.

Assuming gifts must be non-negative, $g^*$ will always be greater than $g'$. This is not enough to show that inefficient concessions are sometimes optimal. To show this, consider high type equilibrium utility and assume that the high types are perfectly patient (that is, $\de_h=1$):  
\be
U_h=pT(1+eg)+peg+(1-p)(W-D(1+eg))\footnote{Note the gift payment does not enter here: the current period is weighted $1-\de$, which is 0, while the future is weighted $\de=1$.}
\ee

Whether it is beneficial for high types to settle on the efficient concession or the inefficient concession needs to be determined by comparing the utility under each. For low $p$, the term $(1-p)(W-D(1+eg))$ dominates, showing that clearly, as low an $e$ as possible is best, so $e=0$, that is, complete inefficiency is the best possible equilibrium efficiency level. It is thus rational under such parameters for inefficient gifts to be given as concessions. \hfill $\blacksquare$

@@

\begin{theorem}
	Assume that concessions have future material value.
	\begin{enumerate}
		\item[(a)] If future material value cannot be destroyed, a mediator can be used to expanded the parameter space over which peace can be achieved.
		\item[(b)] If future material value can be destroyed, a mediator can be used to eliminate inefficient concessions and improve welfare.
	\end{enumerate}
	\label{theorem:5}
\end{theorem}
\emph{Proof:}  The low type IC ($pg_h +X_{FF}^l \geq pX_{FT}^l+(1-p)X_{FF}^l-g_h+pg_h$) binds, so the equilibrium benchmark gift is $g^{*}=p(X_{FT}^l-X_{FF}^l)$ and high type utility is $U_h^*= pX_{TT}+(1-p)X_{FF} - (1-p)g^*$). Under $M$, the low type IC can bind without having high types give a concession to low types. The equilibrium separating gift $g^{*M}$ will be larger than $g^*$ in this case, specifically 
\be
g^{*M}= \frac{p(X_{FT}^l-X_{FF}^l)}{1-p}
\ee
High type utility under $M$ is 
\be
U_h^M= pX_{TT}+(1-p)X_{FF}
\ee
because high types exchange concessions of the same value. $U_h^M>U_h^*$. Note that no additional wealth is created by $M$, but less is transferred to low types, and welfare is defined as high-type utility.\footnote{Low types IR constraint still binds and $U_l=X_{FF}$} \hfill $\blacksquare$

@@

>>>>

\section{Analysis of Benchmark Cases}
\label{sec:bench}
When countries are faced with doubts about each other's motives, they have several options of how to proceed. There are three basic types of equilibria that this analysis will examine.\footnote{If the discount factor for the high type ($\delta_h$) exceeds the cutoff found in expression (1), there are many types of equilibria; consider for instance oscillating every other period between Peace and War with a grim trigger War threat. In this paper, we are not interested in these equilibria because they are not as realistic for applications to conflict scenarios. These are also always payoff dominated by at least one of the other equilibria discussed, so are not attractive.} A thorough analysis of the different types of equilibria, as well as all proofs for this section, can be found in the appendix. 
\\
\\
{\bf \emph{Pooling equilibrium}\\}
\vskip-.2in
In the \emph{pooling equilibrium}, both countries do not give concessions in period 0, then always play Fight in every period beginning with period 1. 
Countries continue to doubt, doing nothing to bridge the credibility gap, and thus not cooperating. This is a \emph{pooling equilibrium}, and is likely to happen when countries are in long-set patterns of distrust. Active fighting is not necessary; remaining outside of a state of peace is all that is required.  See Lemma~\ref{lemma:1} and Lemma~\ref{lemma:2} in the appendix for a formal characterization of this type of equilibrium.

@@

The patience threshold to separate through concessions is much higher under these parameters when the concessions carry future material value. In order to calculate it, we need to know the size of the minimum separating concession, which is again calculated from the low-type incentive constraint. The minimum separating concession is $g^1 = \frac{p(1-d_l)(D+T)}{(1-d_l)(1-pT)+(1-p)D} = .\overline{63}$.\footnote{See proof of Theorem~\ref{theorem:4} below for details.} The minimum patience level required for separation is $\de_h^1 = D(1-p)+1-pT + \frac{p(W-(T+D))}{g^1} = .7\overline{9}.$ Because the specified $\de_h$ is not involved in any of these calculations, we can see that for any $\de_h \in [.5\overline{3},.7\overline{9})$, separating through concessions is possible when concessions have no material value but is not possible when they do. \hfill $\blacksquare$

@@

\begin{theorem}
		Assume the high types are not patient enough to separate without concessions and that concessions have material value that can be destroyed. There are parameters under which the optimal equilibrium is a separating equilibrium in which concessions are inefficient.
	\label{theorem:4}
\end{theorem}

\emph{Proof}: See appendix.

Inefficient concessions are optimal under low $p$ (there's a good chance of facing a low type who will use the concession against you) as well as when $\de_l$ is low ($g^*$ is decreasing in the patience of the low type, as a larger concession is required to deter an impatient low-type from being untruthful).

@@

\begin{theorem}
	Assume that concessions have future material value.
	\begin{enumerate}
		\item[(a)] If future material value cannot be destroyed, a mediator can be used to expanded the parameter space over which peace can be achieved.
		\item[(b)] If future material value can be destroyed, a mediator can be used to eliminate inefficient concessions and improve welfare.
	\end{enumerate}
	\label{theorem:5}
\end{theorem}
\emph{Proof:}  The low type IC ($peg_h +X_{FF}^l \geq pX_{FT}^l+(1-p)X_{FF}^l-g_h+peg_h$) binds, so the equilibrium benchmark gift is $g^{*}=p(X_{FT}^l-X_{FF}^l)$ and high type utility is $U_h^*= pX_{TT}+(1-p)X_{FF} - (1-p)g^*$). Under $M$, the low type IC can bind without having high types give a concession to low types. The equilibrium separating gift $g^{*M}$ will be larger than $g^*$ in this case, specifically 
\be
g^{*M}= \frac{p(X_{FT}^l-X_{FF}^l)}{1-p}
\ee
High type utility under $M$ is 
\be
U_h^M= pX_{TT}+(1-p)X_{FF}
\ee
because high types exchange concessions of the same value. $U_h^M>U_h^*$. Note that no additional wealth is created by $M$, but less is transferred to low types, and welfare is defined as high-type utility.\footnote{Low types IR constraint still binds and $U_l=X_{FF}$} \hfill $\blacksquare$

@@

\emph{Proof of Theorem \ref{theorem:4}:}\\
Because it is relatively complicated to calculate equilibrium concessions in the general case, a slight simplification of the general model will be made for proof of inefficient concessions. Restrict functional forms of $T(\cdot),W(\cdot)$ and $D(\cdot)$ so that each is a scalar multiplied by its arguments. Also assume that $s_1=s_2=m_1=m_2=1.$ Because of linearity, equilibrium optimization has high types choosing $\alpha=1$ (because a concession would only come from another high type in equilibrium) and low types choosing $\alpha=0$. Note that none of these simplifications is essential for the results to hold but we make them for ease of calculation and demonstration of the results. The low type's IC constraint binds so with $e=1$, $g^*$ can be calculated to be $g^*= \frac{p(1-d_l)(D+T)}{(1-d_l)(1-pT)+(1-p)D}$. The high type's utility when $e=1$ is $\frac{1}{1-d_h}\left\{pT(1+g^*) + (1-p)(W-D(1+g^*))\right\} - g^*$.

When $e=0$---that is, the future material value of the concessions is destroyed---we are back in the case of Section~\ref{sec:bench} and the minimum separating concession is $g'=p(T+D)$. The threshold for high types to cooperate is again $\frac{W}{T+D}$. High type utility when $e=0$ is $\frac{1}{1-d_h}\left\{pT + (1-p)(W-D)\right\} - g'$.

Whether it is beneficial for high types to settle on the efficient concession or the inefficient concession needs to be determined by comparing the utility under each, and either can be larger. Inefficient concessions are made optimal when $p$ is low (there's a good chance of facing a low type who will use the concession against you) as well as when $\de_l$ is low ($g^*$ is decreasing in the patience of the low type, as a larger concession is required to deter an impatient low-type from being untruthful). \hfill $\blacksquare$

### 2017-09-12 notes.tex
new proof and exposition around theorem 4

\newpage
To-do list after Venice CesIfo 06/15/17:
\begin{itemize}
	\item From Eli four years ago: why can mediator do what the two sides cannot?
	\item Add discussion of intuitive criterion to text (if I end up needing it): out of eqm beliefs put zero weight on types that can never gain from deviating from a fixed eqm outcome
	\item Incorporate mediator result into text; re-organize text so it matches June presentation (re-worded Theorem 4, new Theorem 5)
		\begin{itemize}
			\item Careful of SWOC being chosen over STC ($\de_h$ threshold restriction?)
		\end{itemize}
	\item Note for mediator result:
		\begin{itemize}
			\item For two high types to cooperate in the mechanism, need large enough gift to screen out low types. They're still just reporting, so don't give a gift if not matched with another high type. PLUS when they're exchanging gifts, the gifts will actually be used for good:
				\[
				  T(1+g) \geq (1-\de)[T(1+g) + W] + \de (W-D)
				\]
				\[
				  \de \geq \frac{W}{T + Tg + D}
				\]
				Note that this threshold is smaller than $\frac{W}{T + D}$, the one we get without concessions having material benefit / harm
		\end{itemize}
	\item Jim Fearon: maybe get rid of repeated game and just parameterize a one-shot game (doens't work because concessions required for separation are too large; require future to recoup cost
		\begin{itemize}
			\item he has a paper about concessions being used against you; he also gave me the reference for another one but I've forgotten it
		\end{itemize}
	\item Someone who \textit{can't} cooperate under either $e=0$ or $e=1$ CAN cooperate under mediator
		\begin{itemize}
			\item In the absence of mediation, being able to burn money can help, but you still lose something
			\item I still have a question about the overall game: if $e=0$ provides higher welfare if it were enforceable but it's not, do we just go with $e=1$?
			\item Some parameters. For all of these, I take $W=8$, $D=5$, $T=10$
				\begin{itemize}
					\item $p=.4, \de_l = .45, \de_h = \de_{STC} = .53$. Choose $e=1$, but $\de_1 = .65$
					\item $p=.3, \de_l = .45, \de_h = \de_{STC}  = .53$. Choose $e=1$, but $\de_1 > 1$; even worse at $p=.2$
					\item $p=.4, \de_l = .5, \de_h = \de_{STC} = .53$. Choose $e=1$, $\de_1 \geq .4$ (works fine)
					\item $p=.1, \de_l = .1, \de_h = \de_{STC} = .53$. Choose $e=0$, but $\de_0 \geq 6.7$
					\item $p=.05, \de_l = .5, \de_h = .7$. Choose $e=0$, but $\de_0 = 7.6$
					\item $p=.1, \de_l = .5, \de_h = .75$. Choose $e=0$, but $\de_0 = 6.7$ and $\de_1 \geq 3$
				\end{itemize}
		\end{itemize}

>>>>

\newpage
To-do list after Venice CesIfo 06/15/17:
\begin{itemize}
	\item From Eli four years ago: why can mediator do what the two sides cannot?
	\item Add discussion of intuitive criterion to text (if I end up needing it): out of eqm beliefs put zero weight on types that can never gain from deviating from a fixed eqm outcome
	\item Incorporate mediator result into text; re-organize text so it matches June presentation (re-worded Theorem 4, new Theorem 5)
		\begin{itemize}
			\item Careful of SWOC being chosen over STC ($\de_h$ threshold restriction?)
		\end{itemize}
	\item Note for mediator result:
		\begin{itemize}
			\item For two high types to cooperate in the mechanism, need large enough gift to screen out low types. They're still just reporting, so don't give a gift if not matched with another high type. PLUS when they're exchanging gifts, the gifts will actually be used for good:
				\[
				  T(1+g) \geq (1-\de)[T(1+g) + W] + \de (W-D)
				\]
				\[
				  \de \geq \frac{W}{T + Tg + D}
				\]
				Note that this threshold is smaller than $\frac{W}{T + D}$, the one we get without concessions having material benefit / harm
		\end{itemize}
	\item Jim Fearon: maybe get rid of repeated game and just parameterize a one-shot game (doens't work because concessions required for separation are too large; require future to recoup cost)
		\begin{itemize}
			\item he has a paper about concessions being used against you; he also gave me the reference for another one but I've forgotten it
		\end{itemize}
	\item Someone who \textit{can't} cooperate under either $e=0$ or $e=1$ CAN cooperate under mediator
		\begin{itemize}
			\item In the absence of mediation, being able to burn money can help, but you still lose something
			\item Some parameters. For all of these, I take $W=8$, $D=5$, $T=10$
				\begin{itemize}
					\item $p=.4, \de_l = .45, \de_h = \de_{STC} = .53$. Choose $e=1$, but $\de_1 = .65$
					\item $p=.3, \de_l = .45, \de_h = \de_{STC}  = .53$. Choose $e=1$, but $\de_1 > 1$; even worse at $p=.2$
					\item $p=.4, \de_l = .5, \de_h = \de_{STC} = .53$. Choose $e=1$, $\de_1 \geq .4$ (works fine)
					\item $p=.1, \de_l = .1, \de_h = \de_{STC} = .53$. Choose $e=0$, but $\de_0 \geq 6.7$
					\item $p=.05, \de_l = .5, \de_h = .7$. Choose $e=0$, but $\de_0 = 7.6$
					\item $p=.1, \de_l = .5, \de_h = .75$. Choose $e=0$, but $\de_0 = 6.7$ and $\de_1 \geq 3$
				\end{itemize}
		\end{itemize}
